---
title: "Visualize Elementary Flux Modes from E. coli model"
author: "Chaitra Sarathy"
date: "July 11, 2019"
output: html_document
---
This code is part of the analysis used in the study: 
"An integrative workflow to visualize Elementary Flux Modes in genome-scale metabolic models", Sarathy et al. (2019) (in preparation)

It performs step D of the workflow described in "tutorial_efmviz_ecoli.mlx", i.e., visualizes the EFMs that were selected after running the tutorial. In order to visualize an EFM and map fluxes on reactions, follow the steps below.


First, the the SBML file(s) created in the mlx script are loaded into Cytoscape. 
After which a set of functions from RCy3 are called to perform network layout operations and mapping reaction fluxes on the visualized network. 
This automation allows users to visualize multiple EFMs in shorter time. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

DEPENDENCIES
- R libraries: RCy3
- Cytpscape (v 3.5 and above)
- Cytpscape apps: cy3sbml, yFiles Layout

INPUTS
- an EFM/submodel saved as an SBML file (as generated in the tutorial)
- file containing reaction fluxes (.txt): two-column comma-separated file <reaction ID><flux value> (as generated in the tutorial)
- Network style file provided under core/inputs/efmviz_style.xml

EXPECTED OUTCOME
EFM from E. coli that was selected from the tutorial is visualized as a network of genes, reactions and metabolites with reaction fluxes mapped on the edges in the network.

NOTE: Due to current unavailability of API for cy3sbml, unfortunately, the file import must be done manually through the GUI. Also, due to licensing issues, yFiles layout cannot be applied through the script and must be done in the GUI. We have tried our best to automate the network operations wherever possible. 

STEPS (* indicates manual operation in Cytoscape GUI)
A. File import *
B. Set input file names
C. Install and load necessary R packages and connect to Cytoscape
D. Perform network operations and data mapping
  i. Import and apply the provided style file 
  ii. Color reaction nodes as green
  iii. Get node table of the network to modify node display labels 
  iv. Hide genes and GPR relation nodes
   v. Map reaction fluxes
E. Apply layout*
-----------------------------------------------------------------------------------------------------------------------------------------------------------------------
A. File import*

i. Open Cytoscape (v 3.5 and above) and make sure cy3sbml app has been installed. If installing for the first time, a Cytoscape restart maybe necessary
ii. Click on the cy3sbl app-specific icon and select the EFM (SBML submodel file which was generated from MATLAB) to open
Three networks are generated for every file imported and they are respectively identified using the prefixes: All, Base and Kinetic. 
iii. Click on the Base network and run the script below. (Other two networks, All and Kinetic may be deleted from GUI)
-----------------------------------------------------------------------------------------------------------------------------------------------------------------------
B. Before running this script, declare file locations and file names. 
-----------------------------------------------------------------------------------------------------------------------------------------------------------------------
NOTE: CHANGE THE NAMES OF ONLY THESE VARIABLES
```{r}
dirname(rstudioapi::getActiveDocumentContext()$path)
usecase1_fileName = "ecolidataForMapping_EFM_885.txt"
styleName = "efmviz_style"
```
C. Install and load necessary R packages
```{r}
if("RCy3" %in% rownames(installed.packages()) == FALSE) {install.packages("RCy3")}
if("readxl" %in% rownames(installed.packages()) == FALSE) {install.packages("readxl")}
if("RColorBrewer" %in% rownames(installed.packages()) == FALSE) {install.packages("RColorBrewer")}
if("rstudioapi" %in% rownames(installed.packages()) == FALSE) {install.packages("rstudioapi")}

library(RCy3)
library(readxl)
library(RColorBrewer)
library(rstudioapi)
```

Make sure the connection between R and Cytoscape works
```{r}
cytoscapePing() #must return "You are connected to Cytoscape!"
```
cytoscapePing() must return "You are connected to Cytoscape!". If not, Cytoscape needs a restart.
-----------------------------------------------------------------------------------------------------------------------------------------------------------------------
NOTE: MAKE SURE THE BASE NETWORK IS SELECTED IN CYTOSCAPE

D. Perform network operations and data mapping

i. Import and apply the provided style file 
```{r}
setVisualStyle(style.name = styleName)
```
ii. Color reaction nodes as green
```{r}
clearSelection()
setNodeColorBypass(node.names = getSelectedNodes(selectNodes(nodes = "reaction", by.col = "sbml type")), new.colors = "#00FF00" )
```
iii. Get node table of the network to modify node display labels 
```{r}
# Get four columns label (default value of label), name (node identifier), sbml type (for filtering nodes later) and cyId (will be used for assigning identifiers for edges)
nodeTab = getTableColumns(table = "node", columns = c("label", "name","sbml type", "cyId"))
nodeTab[,"nodeLabels"] = character()
nodeTab$nodeLabels = nodeTab$label

# Remove prefix
nodeTab$commonId = nodeTab$cyId
nodeTab$commonId = gsub("^R_", "",  nodeTab$commonId)
nodeTab$commonId = gsub("^M_", "",  nodeTab$commonId)
nodeTab$commonId = gsub("^G_", "",  nodeTab$commonId)

# Remove reaction label
nodeTab$nodeLabels[which(nodeTab$`sbml type` == "reaction")] = ""

# Load the updated node table into Cytoscape 

loadTableData(data = nodeTab, data.key.column = "cyId", table = "node", table.key.column = "cyId")

# Remap node labels to new label
updateStyleMapping(style.name = styleName, mapVisualProperty('node label', 'nodeLabels','p'))
```

iv.  Hide genes and GPR relation nodes

```{r}
clearSelection()
hideNodes(node.names = getSelectedNodes(selectNodes(nodes = "fbc_geneProduct", by.col = "sbml type")))
clearSelection()
hideNodes(node.names = getSelectedNodes(selectNodes(nodes = "fbc_and", by.col = "sbml type")))
clearSelection()
hideNodes(node.names = getSelectedNodes(selectNodes(nodes = "fbc_or", by.col = "sbml type")))
clearSelection()
```

v. Map reaction fluxes

```{r}
# Load reaction flux
reactionFlux <- read.csv(file=usecase1_fileName)
```
Extract the edge table and assign flux value to each edge
In this network, the edges did not have any unique identifiers. 
So, identifier of each reaction node was copied onto all the connected reactant and product nodes using the section below.

```{r}
# Get the edge table
edgeTab = getTableColumns(table = "edge", columns = c("interaction type"))
edgeTab$edgeID = rownames(edgeTab)
edgeTab[,"edgeName"] = character()

# Get all reaction nodes
rxnNodes = getSelectedNodes(selectNodes(nodes = "reaction", by.col = "sbml type"))
clearSelection()

# for each reaction node, get associated edges
# for each edge if interaction type is is reaction-product or reaction-reaction, add the reaction id as edge identifier

# NOTE: This for loop will take long time for big networks    

for (ii in rxnNodes){
  rxnEdges = selectEdgesAdjacentToSelectedNodes(selectNodes(nodes = ii, by.col = "shared name"))$edges
  for (jj in rxnEdges){
    ind = which(edgeTab$edgeID == jj)
    if (edgeTab$`interaction type`[ind] == "reaction-product" || edgeTab$`interaction type`[ind] == "reaction-reactant"){
      edgeTab$edgeName[ind] = nodeTab$commonId[match(ii, nodeTab$name)]
    }
  }
  clearSelection()
}
# The prefix was removed since the data file did not contain the same prefix for reaction IDs 
# edgeTab$edgeName = gsub("^R_", "",  edgeTab$edgeName)

# Make sure flux data in edge table is numeric
edgeTab[,"fluxData"] = numeric()
edgeTab$fluxData = reactionFlux$Flux[match(edgeTab$edgeName, reactionFlux$Reactions)]
edgeTab[which(is.na(edgeTab$fluxData)), 4] = 0
edgeTab$fluxData <- signif(edgeTab$fluxData,3)

# Load the modified edge table, now containing edge IDs and flux values to Cytoscape
loadTableData(data = edgeTab, data.key.column = "edgeID", table = "edge", table.key.column = "SUID")

# Map the flux data on the edges, set edge color and edge weight
setEdgeLineWidthMapping('fluxData', c(0,30), c(2,15), style.name=styleName)
setEdgeColorDefault("#969696", style.name=styleName)
setEdgeColorMapping('fluxData', c(0,1,30), c("#999999","#FFCCCC","#FF0000"), style.name=styleName)

# Map edge label to the flux value
setEdgeLabelMapping('fluxData', style.name = "efmviz_style")
```
-----------------------------------------------------------------------------------------------------------------------------------------------------------------------
E. Apply layout*
To apply layout, click Layout --> yFiles Orthogonal Layout

```{r}
# Additional section (Run below code for EFM backbone networks to display EFM occurrence % on the reaction nodes)
# nodeTab = getTableColumns(table = "node", columns = c("cyId","sbml type","efmOccPercentage","nodeLabels"))
# nodeTab$efmOccPercentage = as.character(signif(nodeTab$efmOccPercentage, 3))
# nodeTab$nodeLabels[which(nodeTab$`sbml type` == "reaction")] = paste(nodeTab$efmOccPercentage[!is.na(nodeTab$efmOccPercentage)], "%")
# loadTableData(data = nodeTab, data.key.column = "cyId", table = "node", table.key.column = "cyId")
# updateStyleMapping(style.name = styleName, mapVisualProperty('node label', 'nodeLabels','p'))
```